Polynomial Time and Full Dependent Types

### 2023-07-09 paper/paper.tex
[ paper ] Lots of writing

@@

\newcommand{\Let}{\mathrm{let}}
\newcommand{\LetPair}{\mathrm{letpair}}
\newcommand{\In}{\mathrm{in}}
\newcommand{\If}{\mathrm{if}}
\newcommand{\dupNat}{\mathrm{dupNat}}

\newcommand{\istype}{\mathrm{type}}
\newcommand{\isctxt}{\mathrm{ctxt}}

@@

\newcommand{\LinPoset}{\mathrm{LinPoset}}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Linear $\lambda$-Calculus and Implicit Computational
  Complexity}
\label{sec:linear-icc}

Introduce the idea of linear typing for complexity control.

Linear $\lambda$-calculus, as we shall be discussing it in this
section:
\begin{mathpar}
  \inferrule* [right=Var]
  { }
  {x : A \vdash x : A}

  \inferrule* [right=Lam]
  {\Gamma, x : A \vdash M : B}
  {\Gamma \vdash \lambda x.M : A \multimap B}

  \inferrule* [right=App]
  {\Gamma_1 \vdash M : A \multimap B \\
    \Gamma_2 \vdash N : A}
  {\Gamma_1, \Gamma_2 \vdash M\,N : B}
\end{mathpar}

\begin{mathpar}
  \inferrule* [right=$\otimes$-Intro]
  {\Gamma_1 \vdash M : A \\ \Gamma_2 \vdash N : B}
  {\Gamma_1, \Gamma_2 \vdash (M,N) : A \otimes B}

  \inferrule* [right=$\otimes$-Elim]
  {\Gamma_1 \vdash M : A \otimes B \\ \Gamma_2, x : A, y : B \vdash N : C}
  {\Gamma_1, \Gamma_2 \vdash \Let\,(x,y)=M\,\In\,N : C}
\end{mathpar}

\subsection{Excluding Recursion}

What do we get if we exclude recursion altogether?

What happens if we include natural number recursion naively?

\paragraph{A Binary Tape Datatype}

FIXME: how to do this?

move left, move right, update. Tape is always finite.

\subsection{Cons-free programming}

FIXME: zero case has a context, or do we $\lambda$-abstract?
\begin{displaymath}
  \inferrule*
  {\vdash M_z : A \\ x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero \mapsto M_z; \conSucc(x) \mapsto M_s\} : A}
\end{displaymath}

\begin{displaymath}
  \inferrule*
  {\Gamma \vdash M : \tyNat}
  {\Gamma \vdash \dupNat\,M : \tyNat \otimes \tyNat}
\end{displaymath}

Intuitive reason it works: we can only iterate over the natural
numbers we have recieved as input, there is no way to construct new
natural numbers by iteration. Duplicating natural numbers is safe,
becasue we can only do it a number of times controlled by iteration of
the natural numbers we are given.

FIXME: is it possible to pin down how it relates to SAL?

\paragraph{Completeness}

\begin{displaymath}
  \begin{array}{l}
    I_1 : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_1 = \lambda n. \lambda s.\tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto f\,s \}
  \end{array}
\end{displaymath}

\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_{k+1} = \lambda n. \lambda s.
    \begin{array}[t]{@{}l}
      \Let\,(n,n') = \dupNat\,n\,\In\\
      \tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto I_k\,n'\,s \}
    \end{array}
  \end{array}
\end{displaymath}

@@

The cons-free style of programming allows us to express all polytime
Turing machines, but is quite awkward from the point of view of
functional programming. It allows us to iterate over natural numbers
that come from the input but does not allow us to build further values
to do iteration on. For example, if our input is a list, then we
cannot transform it into a binary search tree and then flatten it
\bob{actually write this}.

Introduce a new type $\Diamond$. Each $\Diamond$ value represents a
``chunk of iterability''.

{\bf Rough model:} programs are polynomials, data are bags of
diamonds. Applying a program to some data yields a concrete number of
steps that will bound the resulting computation.

\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M : \Diamond}
  {\Gamma \vdash \conZero(M) : \tyNat}

  \inferrule*
  {\Gamma_1 \vdash M : \Diamond \\ \Gamma_2 \vdash N : \tyNat}
  {\Gamma_1, \Gamma_2 \vdash \conSucc(M, N) : \tyNat}
\end{mathpar}

\begin{displaymath}
  \inferrule*
  {d : \Diamond \vdash M_z : A \\ d : \Diamond, x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero(d) \mapsto M_z; \conSucc(d,x) \mapsto M_s \} : A}
\end{displaymath}

\paragraph{Other datatypes} Trees, and so on.

\paragraph{Completeness for Polytime} We now show how every polytime
Turing machine can be simulated in LFPL with a non-iterable type of
tapes. The construction, due to Aelhig and Schwictenberg
\cite{syntactic-lfpl}, illustrates how programming with diamonds in
LFPL differs from the soft system in the previous section.

Assume a state type $\mathit{St}$ and a step function $f : \mathit{St} \multimap \mathit{St}$.

The linear iterator:
\begin{displaymath}
  \begin{array}{l}
    I_1 : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_1 = \lambda (n, s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d), s);\\
      \conSucc(d,(n,s)) &\mapsto& (\conSucc(d,n),f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}

Step:\textbf{Is the extra $f\,s$ required here?}
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_{k+1} = \lambda (n,s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d),s); \\
      \conSucc(d,(n,s)) &\mapsto& \Let\,(n,s)=I_k\,(n,s)\,\In\,(\conSucc(d,n), f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}

\begin{lemma}
  (Under a plausible operational semantics) FIXME: use an equational
  theory instead, that includes $\beta$-equality.
  \begin{displaymath}
    I_k\,(\underline{n},s) \longrightarrow^* (\underline{n},v)
  \end{displaymath}
  iff
  \begin{displaymath}
    f^{\binom{n}{k}}\,s \longrightarrow^* v
  \end{displaymath}
\end{lemma}

\subsection{Explicit Resource Tracking}

\begin{enumerate}
\item We still require diamonds to build zero and successor, but the
  iterator does not give them back.
\item In the simply typed case, we can annotate types with the static
  amount of potential they have per constructor.
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Complexity Control in Quantitative Type Theory}
\label{sec:qtt}

We have now seen three systems for controlling resource usage by means
of linear typing. We now look to extend these systems to include
dependent types. Our motivation is threefold:
\begin{enumerate}
\item We would like to have a means for reasoning about the behaviour
  of the programs we write in our systems. FIXME: why dependent types
  then? Why not a program logic?
\item In case of the explicitly resource accounted system, simple
  typing is limited in the reasoning the programmer can perform about
  the resources required by a particular program. While we could in
  principle
\item As well as constructing proofs \emph{about} complexity
  constrained programs, we would also be able to write constructive
  proofs that are themselves of constrained complexity.
\end{enumerate}

To integrate linear typing for controlling time complexity with
dependent types we use \emph{Quantitative Type Theory} (QTT)
\cite{atkey18qtt}. In this section, we review the definition of QTT
and how we have adapted it to implicit and explicit control of
computational complexity.

\begin{enumerate}
\item General description of QTT
\item Non-iterable datatypes in QTT
\item Adapting QTT to include a ``soft'' natural number type
\item Diamonds
\item
\end{enumerate}

@@

\begin{enumerate}
\item Rules for contexts (using $\epsilon$ for empty contexts to avoid
  confusion between $\diamond$ and $\Diamond$):
  \begin{mathpar}
    \inferrule* [right=Ctxt-Emp]
    { }
    {\epsilon~\isctxt}

@@

    \inferrule* [right=Tm-Conv]
    {\Gamma \vdash M \stackrel\sigma: S \\ 0\Gamma \vdash S \equiv T~\istype}
    {\Gamma \vdash M \stackrel\sigma: T}
  \end{mathpar}
\item Rules for $\Pi$ and $\Sigma$ types
  \begin{mathpar}
    \inferrule* [right=$\Pi$-type]
    {0\Gamma \vdash S~\istype \\ 0\Gamma, x \stackrel0: S \vdash T~\istype}
    {0\Gamma \vdash (x \stackrel\rho: S) \to T~\istype}

@@

  FIXME: $\Sigma$-types
\item Rules for equality
  \begin{mathpar}
    \inferrule* [right=Eq-type]
    {0\Gamma \vdash S~\istype \\
      0\Gamma \vdash M \stackrel0: S \\
      0\Gamma \vdash N \stackrel0: S}
    {0\Gamma \vdash M =_S N~\istype}

    \inferrule* [right=Eq-Intro]
    {\Gamma \vdash M \stackrel\sigma: S}
    {\Gamma \vdash \conRefl(M) \stackrel\sigma: M =_S M}
  \end{mathpar}
  \bob{Equality reflection?}
\item Rules for universe
\end{enumerate}

\subsection{Non-iterable Datatypes}

\subsection{Cons-free Natural Numbers}

Of course, ``cons-free'' only applies to the runtime portion of the
system. In the $\sigma = 0$ fragment, we are allowed to create new
natural numbers arbitrarily:
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash}
  {\Gamma \vdash \conZero \stackrel0: \tyNat}

  \inferrule*
  {\Gamma \vdash M \stackrel0: \tyNat}
  {\Gamma \vdash \conSucc(M) \stackrel0: \tyNat}
\end{mathpar}
Creating numbers in the $\sigma = 0$ fragment ensures that we can talk
about natural numbers in our types, allowing us to state properties of
our programs.\bob{Forward ref to actually doing this.}

The eliminator:
\begin{displaymath}
  \mprset{flushleft}
  \inferrule*
  {0\Gamma, m \stackrel0: \tyNat \vdash X~\istype \\\\
    0\Gamma \vdash M_z \stackrel\sigma: X[\conZero/m] \\\\
    0\Gamma, n \stackrel0: \tyNat, x \stackrel1: X[n/m] \vdash M_s \stackrel\sigma: X[\conSucc(n)/m] \\\\
    \Gamma \vdash N \stackrel\sigma: \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero \mapsto M_z; \conSucc(n/x) \mapsto M_s\} \stackrel\sigma: X[N/m]}
\end{displaymath}

\subsection{Diamonds}

\begin{mathpar}
  \inferrule* [right=DiaType]
  {\Gamma \vdash}
  {0\Gamma \vdash \Diamond~\istype}

  \inferrule* [right=DiaIntro]
  {\Gamma \vdash}
  {0\Gamma \vdash * \stackrel0: \Diamond}

  \inferrule* [right=DiaEta]
  {\Gamma \vdash M \stackrel0: \Diamond}
  {\Gamma \vdash M \equiv * \stackrel0: \Diamond}
\end{mathpar}

Link to Selinger et al: the $\eta$ rule for $\Diamond$s means that are
essentially depending on the ``shape'' of anything that depends on
diamonds.

\subsection{LFPL-style Natural Numbers}

\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M \stackrel\sigma: \Diamond}
  {\Gamma \vdash \conZero(M) \stackrel\sigma: \tyNat}

>>>>

\newcommand{\TapeType}{\mathrm{Tape}}

@@

\newcommand{\Let}{\mathrm{let}}
\newcommand{\LetPair}{\mathrm{letpair}}
\newcommand{\In}{\mathrm{in}}
\newcommand{\If}{\mathrm{if}}
\newcommand{\Then}{\mathrm{then}}
\newcommand{\Else}{\mathrm{else}}
\newcommand{\dupNat}{\mathrm{dupNat}}
\newcommand{\Match}{\mathrm{match}}

% QTT stuff
\newcommand{\istype}{\mathrm{type}}
\newcommand{\isctxt}{\mathrm{ctxt}}

\newcommand{\BoolTy}{\mathrm{Bool}}
\newcommand{\cTrue}{\mathsf{true}}
\newcommand{\cFalse}{\mathsf{false}}
\newcommand{\ListTy}{\mathrm{List}}
\newcommand{\cNil}{\mathsf{nil}}
\newcommand{\cCons}{\mathsf{cons}}

\newcommand{\Rtype}{\mathbf{R}}
\newcommand{\rIntro}{\mathbf{R}}
\newcommand{\rElim}{\mathbf{R}^{-1}}

@@

\newcommand{\LinPreorder}{\mathrm{LinPreorder}}

@@

\renewcommand{\sectionautorefname}{Section}
\renewcommand{\subsectionautorefname}{Section} % PS
\renewcommand{\subsubsectionautorefname}{Section}

@@

Related work: only applications of Linear TT so far are erasure (Abel
et al.) and quantum stuff (Selinger et al.), and (sort of) dependency
analysis. This is another one.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Linear $\lambda$-Calculus and Polytime}
\label{sec:linear-intro}

Not long after Girard introduced Linear Logic \cite{girard87}, it was
observed that its resource sensitivity could be turned to describing
computational complexity classes by purely logical means. Typically, a
logical system is described for which the process of reducing a proof
to a normal form (often by cut elimination) is guaranteed to always be
accomplished within a certain complexity bound. Moreover, the system
is usually proven to be complete for the relevant complexity class, by
constructing a simulation of some known representation. Such systems
that characterise polytime include Bounded Linear Logic (BLL), which
uses explicit polynomials in the formulas \cite{bll} and Soft Affine
Logic (SAL) \cite{lafont}, which does not explicitly represent
time information in formulas, but uses a restricted form of Linear
Logic's $\oc$ modality instead. Light Affine Logic (LAL) \cite{lal} is
another ``counting-free'' system for polytime.

Viewing logical systems though the lens of the Curry-Howard
correspondence, the idea arises that one could define functional
programming languages that characterise complexity classes such as
polytime. SAL has been transformed into a programming language for
polytime by \cite{gaboardi}. Hofmann proposed a new programming
language, Linear Functional Programming Language (LFPL), that uses a
novel ``payment'' system to track iteration. In this section, we
review the use of linear types to capture polytime by presenting two
systems, one based on ideas from SALL and the second more explicitly
based on LFPL. With a view to adapting these systems to dependent
types in \autoref{sec:qtt}, we take an approach slightly different to
much of the polytime linear logic literature. We emphasise the use of
explicit datatypes and eliminators, rather than using impredicative
encodings via universal types. Thus, we are closer to Hofmann's
original LFPL (though not later presentations of it
\cite{dallago-hofmann}) than BLL, SALL, or LAL.

For this section, the linear $\lambda$-calculus we will use will have
linear functions and $\otimes$-products:
\begin{mathpar}
  \inferrule*
  { }
  {x : A \vdash x : A}

  \inferrule*
  {\Gamma, x : A \vdash M : B}
  {\Gamma \vdash \lambda x.M : A \multimap B}

  \inferrule*
  {\Gamma_1 \vdash M : A \multimap B \\
    \Gamma_2 \vdash N : A}
  {\Gamma_1, \Gamma_2 \vdash M\,N : B}

  \inferrule*
  {\Gamma_1 \vdash M : A \\ \Gamma_2 \vdash N : B}
  {\Gamma_1, \Gamma_2 \vdash (M,N) : A \otimes B}

  \inferrule*
  {\Gamma_1 \vdash M : A \otimes B \\ \Gamma_2, x : A, y : B \vdash N : C}
  {\Gamma_1, \Gamma_2 \vdash \Let\,(x,y)=M\,\In\,N : C}
\end{mathpar}
These rules are standard, so we do not describe them further except to
note how linear typing uses presence or absences in a context to
control resource usage -- if a variable is in the context it must be
used exactly once. The fact that this discipline interferes with
dependent types is one of the reasons we turn to QTT when we wish to
add dependent types in \autoref{sec:qtt}.

\subsection{No Recursion, Only Case Analysis}

It is not too difficult to see that reduction of linear
$\lambda$-terms always takes a number of steps linearly proprotional
to the size of the term. This is because every $\beta$-redex
substitutes into exactly one variable, reducing the size of the term
by one each time.

We can increase the expressivity, but not the computational
complexity, of the system by adding datatypes that do not allow
iteration. These can be used for representation, but not for driving
computation. We include the rules here, to show how linearity must be
preserved in these rules, and to foreshadow their dependently typed
counterparts in \autoref{sec:qtt-noniter}. The first type is the
booleans, which are anyway non-recursive and so wouldn't allow
iteration anyway:
\begin{mathpar}
  \inferrule*
  { }
  {\vdash \cTrue, \cFalse : \BoolTy}

  \inferrule*
  {\Gamma_1 \vdash M : \BoolTy \\ \Gamma_2 \vdash N_1 : A \\ \Gamma_2 \vdash N_2 : A}
  {\Gamma_1, \Gamma_2 \vdash \If\,M\,\Then\,N_1\,\Else\,N_2 : A}
\end{mathpar}
The if-then-else rule is careful to ensure that the resources used by
the eliminated $\BoolTy$ and the resources used by the chosen branch
are accounted for separately. The two branches must have the same
resource usage.

Construction and case analysis of lists are given by the following
rules:
\begin{mathpar}
  \inferrule*
  { }
  {\vdash \cNil : \ListTy(A)}

  \inferrule*
  {\Gamma_1 \vdash M : A \\ \Gamma_2 \vdash N : \ListTy(A)}
  {\Gamma_1, \Gamma_2 \vdash \cCons(M,N) : \ListTy(A)}

  \inferrule*
  {\Gamma_1 \vdash M : \ListTy(A) \\
    \Gamma_2 \vdash N_1 : B \\
    \Gamma_2, h : A, t : \ListTy(A) \vdash N_2 : B}
  {\Gamma_1, \Gamma_2 \vdash \Match\,M\,\{\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\} : B}
\end{mathpar}
Hence, we can construct lists arbitrarily, but we can only do case
analysis on them. If we wish to explore a list to a arbitrary depth it
must be driven by an iterable type of the form that we will see in the
next two sections.

With booleans and lists, we can construct several other useful
types. For example, to simulate Turing machines, one can construct a
$\mathit{Tape}$ type as a Zipper \cite{huet}
$\ListTy(\BoolTy) \otimes \BoolTy \otimes \ListTy(\BoolTy)$,
representing a position on the tape with the items before, under, and
after the head.

\subsection{The Cons-free system}
\label{sec:cons-free-intro}

Polynomial time is usually seen as a proxy for ``feasible''
computation. On the face of it, there does not seem to be any
particular reason why polynomials have anything to do with
feasibility. However, one can arrive at the definition of polynomial
time in three steps, by assuming that (i) iterating over the whole
input is feasible; (ii) if two computations are feasible, then so is
their composition; and (iii) performing a feasible computation for
every element of the input is also feasible. It is the last point that
allows complexities of arbitrary polynomial degree to be constructed
(we will see this in action in the completeness construction below and
soundness proofs in \autoref{sec:iterator-soundness}).

Following these ideas, let us assume that the input is a natural
number, so we assume that there is some type of natural numbers
$\tyNat$. For point (i), we must be able to iterate over these natural
numbers, so we use a linear iterator defined by this typing rule:
\begin{displaymath}
  \inferrule*
  {\vdash M_z : A \\ x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero \mapsto M_z; \conSucc(x) \mapsto M_s\} : A}
\end{displaymath}
Note that in the zero, $M_z$, and sucessor, $M_s$, cases, the context
is empty to ensure that these cases may be invoked as many times as
required. Point (ii) above is automatically satisfied by being in a
typed $\lambda$-calculus, where it is difficult to stop functions from
being composable. For point (iii), the iterator as given does not
allow us to nest iterations. Once the natural number input $n$ has
been used for an iteration, the linear typing discipine prevents us
from using it again (note the two separate contexts $\Gamma_1$,
$\Gamma_2$ in the rule for application). In order to allow nested
iterations, we add an operator to duplicate numbers:
\begin{displaymath}
  \inferrule*
  {\Gamma \vdash M : \tyNat}
  {\Gamma \vdash \dupNat\,M : \tyNat \otimes \tyNat}
\end{displaymath}
Somewhat surprisingly, this system is now sound and complete for
polynomial time. Crucially, this depends on the two things we have
\emph{not} allowed. First, we have disallowed the construction of new
natural numbers via the $\conZero$ and $\conSucc$
constructors\footnote{Actually, $\conZero$ would be acceptable, as
  well as any constant natural number. It is only unrestricted use of
  $\conSucc$ that is dangerous.}. If we were to permit this, then we
could use iteration over the input to construct addition,
multiplication (by repeated addition), and then exponentials (by
repeated multiplication). We therefore refer to this system as the
\emph{Cons-Free} system.  The second prohibited feature is the ability
to duplicate values of function type, even though we have allowed
duplication of iterable naturals. If we were to allow this, then we
would be able to sneak in a form of constructors for natural numbers
by encoding them as eliminators that duplicate a function for every
$\conSucc$ step.

We will see in \autoref{sec:consfree-sound} that this system is sound
for polytime by a realisability argument. Completeness can be seen
more directly by constructing a function that iterates a function for
a statically known polynomial number of times in the size of the
input. Assume that we have a known polynomial
$p(n) = c_dn^d + \dots + c_0$ of degree $d$ with natural number
coefficients and some single step function
$f : \mathit{St} \multimap \mathit{St}$ over a state type
$\mathit{St}$ that runs to completion for input of size $n$ in $p(n)$
steps. Then, using the iterator above we can iterate $f$ over a
$\tyNat$ representing the size of the input:
\begin{displaymath}
  \begin{array}{l}
    I_1 : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_1 = \lambda n. \lambda s.\tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto f\,s \}
  \end{array}
\end{displaymath}
To achieve higher degrees, we can use $\dupNat$ to nest iterations:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_{k+1} = \lambda n. \lambda s.
    \begin{array}[t]{@{}l}
      \Let\,(n,n') = \dupNat\,n\,\In\\
      \tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto I_k\,n'\,s \}
    \end{array}
  \end{array}
\end{displaymath}
By further use of $\dupNat$ and composition to handle addition of
polynomials, the function $f$ can now be iterated $p(n)$ many times,
where $n$ is the input $\tyNat$. Thus, the \emph{Cons-free} system can
represent all polytime computations.

@@

The Cons-free system is sound and complete for polytime, but is quite
awkward from the point of view of functional programming. It allows us
to iterate over natural numbers that come from the input but does not
allow us to build further values to do iteration on. For example, if
our input is a list, then we cannot transform it into a binary search
tree and then flatten it, we must always refer back to the original
natural number input. Even dividing the input into two halves to be
treated separately is difficult.

A more flexible system was proposed by \cite{hofmann99}. The
\emph{Linear Functional Programming Language} (LFPL) takes a different
approach to the \emph{Cons-free} system. Instead of completely
prohibiting construction of new data, constructions is allowed if it
is paid for by values of new type $\Diamond$ (``diamonds''):
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M : \Diamond}
  {\Gamma \vdash \conZero(M) : \tyNat}

  \inferrule*
  {\Gamma_1 \vdash M : \Diamond \\ \Gamma_2 \vdash N : \tyNat}
  {\Gamma_1, \Gamma_2 \vdash \conSucc(M, N) : \tyNat}
\end{mathpar}
Thus, to construct a $\conZero$, we must have a $\Diamond$ to pay for
it, and likewise, to construct a $\conSucc$ we must have a diamond. We
can think of $\Diamond$s as an unit of iterable data. Iterability is
``saved up'' in data during construction, and released during
iteration. Diamonds cannot be created by a program itself, for the
same reason that constructors were prohibited in the \emph{Cons-free}
system, but they are released from iterable data during iteration. The
LFPL natural number iterator has the following typing rule:
\begin{displaymath}
  \inferrule*
  {d : \Diamond \vdash M_z : A \\ d : \Diamond, x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero(d) \mapsto M_z; \conSucc(d,x) \mapsto M_s \} : A}
\end{displaymath}
The difference with the \emph{Cons-free} iterator above is that the
$\conZero$ and $\conSucc$ cases now both have an addition binding of
type $\Diamond$. This allows some form nesting of iterations: during
an iteration over the input, the program can accumulate $\Diamond$s to
use for iteration over substructures that are smaller than the current
point in the iteration. A construction, due to Aelhig and
Schwictenberg \cite{syntactic-lfpl}, illustrates how this leads to all
polytime computations. As above, we assume a polynomial $p(n)$ and a
step function $f : \mathit{St} \multimap \mathit{St}$ that needs to be
iterated $p(n)$ times. We construct a linear iterator:
\begin{displaymath}
  \begin{array}{l}
    I_1 : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_1 = \lambda (n, s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d), s);\\
      \conSucc(d,(n,s)) &\mapsto& (\conSucc(d,n),f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Note that this iterator returns the natural number input as well as
the new state. LFPL does not allow duplication of iterable inputs, so
we must always reconstruct it if we want to do further
iteration. Addition of polynomials is accomplished by composition of
iterators. To raise the degree, we again use a nesting iterator:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_{k+1} = \lambda (n,s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d),s); \\
      \conSucc(d,(n,s)) &\mapsto& \Let\,(n,s)=I_k\,(n,s)\,\In\,(\conSucc(d,n), f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Unlike in the \emph{Cons-free} system, this iterator does not raise
the degree of the nested iterator directly. Rather the iterator $I_k$
performs $\binom{n}{k}$ iterations. As observed by Aelhig and
Schwictenberg, this is sufficient because the binomials form a basis
for the vector space of all polynomials.

Despite this slightly more involved completeness construction, the
advantage of LFPL is that it is now easy to have arbitrary iterable
datatypes and to transform between them. We need only take the
introduction and elimination rules for any inductive datatype and add
$\Diamond$ premises to the introduction rules and $\Diamond$ bindings
to the eliminators.

% \subsection{Explicit Resource Tracking}

% \begin{enumerate}
% \item We still require diamonds to build zero and successor, but the
%   iterator does not give them back.
% \item In the simply typed case, we can annotate types with the static
%   amount of potential they have per constructor.
% \end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Quantitative Type Theory}
\label{sec:qtt}

We have now seen two systems for controlling resource usage by means
of linear typing. We now look to extend these systems to include
dependent types. Our motivation is threefold:
\begin{enumerate}
\item We would like to have a means for reasoning about the behaviour
  of the programs we write in our systems. FIXME: why dependent types
  then? Why not a program logic?
\item In case of the explicitly resource accounted system, simple
  typing is limited in the reasoning the programmer can perform about
  the resources required by a particular program. While we could in
  principle
\item As well as constructing proofs \emph{about} complexity
  constrained programs, we would also be able to write constructive
  proofs that are themselves of constrained complexity.
\end{enumerate}

To integrate linear typing for controlling time complexity with
dependent types we use \emph{Quantitative Type Theory} (QTT)
\cite{atkey18qtt}. In this section, we review the definition of QTT
and how we have adapted it to the polytime systems we saw in the
previous section.

@@

\bob{Explain the judgements and general idea of QTT, and the admissibility of $0$-ing}

\subsubsection{Contexts, variables, and conversion}
\begin{mathpar}
  \inferrule* [right=Ctxt-Emp]
  { }
  {\epsilon~\isctxt}

@@

  \inferrule* [right=Tm-Conv]
  {\Gamma \vdash M \stackrel\sigma: S \\ 0\Gamma \vdash S \equiv T~\istype}
  {\Gamma \vdash M \stackrel\sigma: T}
\end{mathpar}

\subsubsection{$\Pi$- and $\Sigma$-types}
\begin{mathpar}
  \inferrule* [right=$\Pi$-type]
  {0\Gamma \vdash S~\istype \\ 0\Gamma, x \stackrel0: S \vdash T~\istype}
  {0\Gamma \vdash (x \stackrel\rho: S) \to T~\istype}

@@

\bob{$\Sigma$-types, from the LICS paper}

\subsubsection{The Identity Type}
\begin{mathpar}
  \inferrule* [right=Eq-type]
  {0\Gamma \vdash S~\istype \\
    0\Gamma \vdash M \stackrel0: S \\
    0\Gamma \vdash N \stackrel0: S}
  {0\Gamma \vdash M =_S N~\istype}

  \inferrule* [right=Eq-Intro]
  {\Gamma \vdash M \stackrel\sigma: S}
  {\Gamma \vdash \conRefl(M) \stackrel\sigma: M =_S M}
\end{mathpar}
\bob{Equality reflection}

\subsubsection{Universe}
\bob{Copy from LICS paper}

\subsubsection{Datatypes} QTT, as we have presented it so far, has no
interesting base types to perform computation on. Following our
presentation of the simply typed linear systems in
\autoref{sec:lin-intro}, we add two kinds of datatype to QTT. First,
\autoref{sec:noniter-qtt}, we add non-iterable datatypes that allow
construction and case analysis, but no recursion. In
\autoref{sec:cons-free-qtt} we describe how to extend QTT to be a
dependently typed adaptation of the Cons-free system of
\autoref{sec:cons-free-intro}. In \autoref{sec:lfpl-qtt}, we apply the
same treatment to the LFPL-style system.

\subsection{Non-iterable Datatypes}
\label{sec:noniter-qtt}

\subsubsection{Booleans} The boolean type was already described in
\cite{atkey18qtt}, but we repeat it here for completeness. Booleans
offer no possibility for iteration in any case, but it is useful to
see how their typing rules handle resource distribution before moving
to more complex types.
\begin{mathpar}
  \inferrule*
  {\Gamma~\isctxt}
  {\Gamma \vdash \BoolTy~\istype}

  \inferrule*
  {\Gamma~\isctxt}
  {0\Gamma \vdash \cTrue, \cFalse \stackrel\sigma: \BoolTy}

  \inferrule*
  {0\Gamma_1, x \stackrel0: \BoolTy \vdash P~\istype \\
    \Gamma_1 \vdash M \stackrel\sigma: \BoolTy \\
    \Gamma_2 \vdash N_t \stackrel\sigma: P[\cTrue/x] \\
    \Gamma_2 \vdash N_f \stackrel\sigma: P[\cFalse/x] \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \If_{x.P}\: M \: \Then \: N_t \: \Else \: N_f \stackrel\sigma: P[M/x]}
\end{mathpar}
The introduction rules for booleans both use a $0$-d context,
indicating that construction of boolean values is free. Elimination of
booleans via a dependently typed if-then-else is more subtle with its
resource usage. The boolean to be eliminated must be constructed in a
context $\Gamma_1$, while the two branches are constructed in context
$\Gamma_2$. Since only one of the branches will be used, sharing
resources between the branches is expected.

One might wonder how, since constructing booleans is $0$-cost by their
introduction rules, the $\Gamma_1$ context will ever be non-$0$. This
is resolved by observing that booleans may be the output of processes
that consume time (e.g., the iteration constructs defined below), and
so $\Gamma_1$ will represent a requirement that the necessary resource
is provided.
